---
title: "MHIvsCrimes"
author: "Aileen Li"
date: "2023-03-08"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r lib}
library(tidyverse)
```

## Tidying
It is important to note we did not include Washington D.C as a state.

The median_housing_income dataset is untidy and contains unnecessary info in relation to what we wanted to analyze, so we modified it for our questions.  

We changed the column names in the dataset to be named in sequential order. Then, we deleted all the columns containing standard error data.
```{r tidy}
# find standard error columns and delete them
med_income_base = h08_1_
colnames(med_income_base) = paste(seq_along(med_income_base))
del_col = list()

for (col in 1:ncol(h08_1_)) {
  if (col!=1 & (col-1)%%2==0) {
    del_col= append(del_col, as.character(col))
  }
}
med_income_base = med_income_base[,!names(med_income_base) %in% del_col]
```

We renamed the columns to their respective titles given in the dataset on the fourth row. And then deleted rows containing unimportant information. 
There were two columns containing different info for the year 2013. We tried to find why they were different, in addition as to what the values in the parentheses meant, but there was no documentation for this dataset. Thus, we decided to keep the first column instance of the year 2013 so that we could have no differing data for the same year and state combination that could possibly affect our graphs.

```{r tidy}
# rename columns
colnames(med_income_base) = med_income_base[4,]

# remove undesired data
med_income_base = med_income_base[-(1:6),-9]

i = min(which(med_income_base$State=='Wyoming'))
med_income_untidy = med_income_base[1:i,] |> filter(State!="D.C.")
```

We made the median_housing_income dataset tidy by changing the individual year columns to be under the Year variable and their values under column Median_Income. Then, we filtered for observations within our desired time period, which is from 1992 to 2016.

```{r tidy}
# make desired data tidy
med_income_tidy = med_income_untidy |> 
  pivot_longer(!State, names_to="Year", values_to="Median_Income") |> 
  filter(substr(Year, 1, 4) %in% (1992:2016))
med_income_tidy$State = toupper(med_income_tidy$State)
```

We checked for years with some values in parentheses and for duplicate years for the same state.
We also removed the values in parentheses for the rest of the years. And for the sake of merging the datasets easier, we completely capitalized the state names.

```{r tidy}
# find years with () and duplicate years other than 2013
pyears = med_income_tidy[grep(")$",med_income_tidy$Year),]
dupl_years = pyears |>
  mutate(compare=substr(Year,1,6)) |>
  select(State, compare) |> group_by(State, compare) |>
  filter(duplicated(compare))

# remove () from some years
med_income_tidy$Year = substr(med_income_tidy$Year, 1,4)
med_income = med_income_tidy
```

## Joining/Merging

```{r}





```



There are 1,300 observations of the 50 states from the year 1992 to 2016. We noticed that there are years with numbers in parentheses, an indication of some kind, but we were unable to determine what these values meant because of the lack of documentation for this dataset. 
